---
title: "18SPostProcessing"
author: "Alexander Evan Chrostowski"
date: "2026-06-15"
output: html_document
---
# Loading packages
```{r}
library(ggplot2); packageVersion("ggplot2")
theme_set(theme_bw())
library(dplyr); packageVersion("dplyr")
library(tidyverse); packageVersion("tidyverse")
#For ordinations and other microbial community analyses
library(phyloseq); packageVersion("phyloseq")
library(vegan); packageVersion("vegan")
library(reshape2); packageVersion("reshape2")
library(randomcoloR); packageVersion("randomcoloR")
library(ggpubr); packageVersion("ggpubr")
library(RColorBrewer)   # for brewer.pal()
library(lubridate)      # for mdy(), year()
```
#Prep of saving results to individual folders for training set comparisons 
```{r}
#TRAINING_SET toggle removed. SILVA_138.2 never worked, it classified nothing
#because LearnTaxa had no rank labels to learn from. DECIPHER is the only
#real option now so this isnt a choice anymore.
out_base <- "/Users/alexchrostowski/Desktop/ALEX - Eggleston Lab /18S_outputs/DECIPHER_18S_213samples"
dir_fig  <- file.path(out_base, "figures")
dir_tab  <- file.path(out_base, "tables")
dir_stat <- file.path(out_base, "stats")
for (d in c(dir_fig, dir_tab, dir_stat)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
cat("exporting to:", out_base, "\n")
```


# Setting all the tables/initial variables
```{r}
#ps18S already built and confirmed earlier today, 15435 taxa, 213 samples,
#E417 to 672. same euro number matching and taxonomy alignment this chunk
#used to do, already done. no rebuild needed, just point everything
#downstream at it.
ps <- ps18S
ps
sum(is.na(tax_table(ps)[,"class"]))
```

# Writing and saving the tables you need for other processing
```{r}
write.table(as(otu_table(ps), "matrix"), file.path(dir_tab, "ASVsF_18S.csv"), sep=",", col.names=NA)
write.table(as(tax_table(ps), "matrix"), file.path(dir_tab, "taxaF_18S.csv"), sep=",", col.names=NA)
write.table(as(sample_data(ps), "matrix"), file.path(dir_tab, "metadata_18S.csv"), sep=",", col.names=NA)
saveRDS(ps, file.path(out_base, "ps_18S.rds"))

# class-frequency table (what's in the bar plots)
class_freq <- as.data.frame(sort(table(tax_table(ps)[, "class"], useNA = "ifany"), decreasing = TRUE))
colnames(class_freq) <- c("class", "n_ASVs")
write.csv(class_freq, file.path(dir_tab, "class_frequency.csv"), row.names = FALSE)
```

# Check which eukaryote groups were recovered
```{r}
#[UPDATED] the eukaryote ranks here are domain, kingdom, phylum, class, order,
#family, genus. no division or supergroup, that was the old broken taxonomy's
#comment left over from before. confirmed columns below, then see what class
#level groups actually showed up.
colnames(tax_table(ps))
sort(table(tax_table(ps)[, "class"], useNA = "ifany"), decreasing = TRUE)
```

# Processing for relative abundance
```{r}
ps_ra <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
sample_data(ps_ra)$Location <- factor(sample_data(ps_ra)$Location, levels=c("MIS","STA","MAB"))
ps_ra_top <- filter_taxa(ps_ra, function(x) sum(x) > .005, TRUE)
ps_ra_top
sorted_dates <- format(as.Date(sample_data(ps_ra_top)$Date, format = "%m/%d/%y"), "%Y-%m-%d")
sample_data(ps_ra_top)$Date <- factor(sorted_dates, levels = sort(unique(sorted_dates)))
all_dates <- as.Date(levels(sample_data(ps_ra_top)$Date))
keep  <- !duplicated(format(all_dates, "%Y-%m"))
month_firsts <- all_dates[keep]
dates_to_show <- format(month_firsts[seq(1, length(month_firsts), by = 3)], "%Y-%m-%d")
dates_to_show
```

```{r}
# each site was sampled on slightly different calendar days, so one shared
# set of quarterly ticks doesn't land inside every panel's actual date range.
# build breaks per site instead so each panel's earliest real date gets a label.
dates_by_site <- sample_data(ps_ra_top) %>%
  as.data.frame() %>%
  group_by(Location) %>%
  distinct(Date) %>%
  mutate(Date = as.Date(Date)) %>%
  arrange(Date) %>%
  filter(!duplicated(format(Date, "%Y-%m"))) %>%
  slice(seq(1, n(), by = 3)) %>%
  pull(Date) %>%
  format("%Y-%m-%d")

dates_by_site
```


# Phylum CHART
```{r}
#[UPDATED] "division" never existed in this taxonomy, that comment was
#leftover from the broken SILVA scheme. phylum is the real rank one step up
#from class here. same data driven palette pattern as before, generated
#fresh so it covers whatever phyla actually showed up in the 213 samples.

#each site was sampled on slightly different calendar days, so one shared
#set of date breaks does not line up with every panel. every 6th month
#instead of every 3rd, so labels have room to not overlap.
dates_by_site <- sample_data(ps_ra_top) %>%
  as.data.frame() %>%
  group_by(Location) %>%
  distinct(Date) %>%
  mutate(Date = as.Date(Date)) %>%
  arrange(Date) %>%
  filter(!duplicated(format(Date, "%Y-%m"))) %>%
  slice(seq(1, n(), by = 6)) %>%
  pull(Date) %>%
  format("%Y-%m-%d")
dates_by_site

phylum_levels <- get_taxa_unique(ps_ra_top, "phylum")
phylum_levels <- phylum_levels[!is.na(phylum_levels)]
palette_p <- setNames(distinctColorPalette(length(phylum_levels)), phylum_levels)

p1_top <- plot_bar(ps_ra_top, x = "Date", fill="phylum")+
  geom_bar(aes(fill=phylum), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16), legend.text = element_text(size=12)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = palette_p, na.value = "grey80") +
  scale_x_discrete(
    breaks = dates_by_site,
    labels = function(x) format(as.Date(x), "%b %Y")
  ) +
  theme(legend.position = "bottom", axis.title = element_text(size=15), title=element_text(size = 25), legend.title = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
p1_top
ggsave(file.path(dir_fig, "phylum_bar_213samples.png"), plot = p1_top, width = 26, height = 8, dpi = 300)
```

# CLASS PLOT
```{r}
# [CHANGED vs 16S] data-driven palette over the classes actually present
# [matches 16S philosophy: hand-curated, hue-grouped, soft tones -- NOT random.
#  related lineages share a hue family. unlisted (rare) classes -> grey80,
#  same as the 16S curated-palette approach.]
euk_class_colors <- c(
  "Chlorophyceae"        = "#4DAF4A",
  "Trebouxiophyceae"     = "#A6D854",
  "Ulvophyceae"          = "#B2DF8A",
  "Mamiellophyceae"      = "#C7E9B4",
  "Chlorodendrophyceae"  = "#D9F0A3",
  "Coleochaetophyceae"   = "#8FD175",
  "Nephroselmidophyceae" = "#6FBF5C",
  "Pedinophyceae"        = "#98D98E",
  "Embryophyta"          = "#3F9142",
  "Chrysophyceae"        = "#E6AB02",
  "Bacillariophyceae"    = "#D9A441",
  "Mediophyceae"         = "#F6C85F",
  "Eustigmatophyceae"    = "#FBE08A",
  "Pelagophyceae"        = "#C9962C",
  "Dinophyceae"          = "#FC8D62",
  "Intramacronucleata"   = "#377EB8",
  "Postciliodesmatophora"= "#80B1D3",
  "Thecofilosea"         = "#984EA3",
  "Glissomonadida"       = "#BC80BD",
  "Imbricatea"           = "#CAB2D6",
  "Perkinsea"            = "#9E6FB0",
  "Cercomonadidae"       = "#B285C7",
  "Phytomyxea"           = "#7D4E96",
  "Vampyrellidae"        = "#A672B8",
  "Chytridiomycetes"     = "#A6761D",
  "Leotiomycetes"        = "#C49A6C",
  "Dothideomycetes"      = "#DACDA0",
  "Sordariomycetes"      = "#8C6432",
  "Eurotiomycetes"       = "#B08248",
  "Entomophthoromycetes" = "#D4B896",
  "Neocallimastigomycetes" = "#946A24",
  "Monogononta"          = "#F4A6C0",
  "Choanoflagellida"     = "#E5C494",
  "Bivalvia"             = "#8DD3C7",
  "Gastropoda"           = "#F2B8D0",
  "Clitellata"           = "#E8909E",
  "Rhabditophora"        = "#F7C9D9",
  "Hydrozoa"             = "#D97590",
  "Demospongiae"         = "#EAA6B8",
  "Phylactolaemata"      = "#F0BFCB",
  "Maxillopoda"          = "#C4909A",
  "Ostracoda"            = "#DFA8B4",
  "Chromadorea"          = "#B87E8A",
  "Actinopterygii"       = "#E88A9E",
  "Mammalia"             = "#D66F84",
  "Aves"                 = "#E0879C",
  "Conoidasida"          = "#5FBFAF",
  "Aconoidasida"         = "#7FC9BC",
  "Ichthyosporea"        = "#B0A8C0",
  "Filasterea"           = "#9F97B3",
  "Incertae Sedis"       = "#A8A0B5",
  "Novel Clade 10"       = "#C4BDD0",
  "Novel Clade 2"        = "#B5AEC5",
  "Novel Clade Gran-4"   = "#A69FBB",
  "Other organisms"      = "grey80"
)

length(euk_class_colors)
setdiff(get_taxa_unique(ps_ra_top, "class"), c(names(euk_class_colors), NA))

c1_top_nolegend <- plot_bar(ps_ra_top, x = "Date", fill="class")+
  geom_bar(aes(fill=class), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = euk_class_colors, na.value = "grey80") +
  scale_x_discrete(
    breaks = dates_by_site,
    labels = function(x) format(as.Date(x), "%b %Y")
  ) +
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 25)) +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
c1_top_nolegend
ggsave(file.path(dir_fig, "class_bar_213samples_NO_LEGEND.png"), plot = c1_top_nolegend, width = 26, height = 10, dpi = 300)

c1_legend_only <- plot_bar(ps_ra_top, x = "Date", fill="class")+
  geom_bar(aes(fill=class), stat="identity",position="stack") +
  scale_fill_manual(values = euk_class_colors, na.value = "grey80") +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4, title = "Class"))

class_legend_grob <- cowplot::get_legend(c1_legend_only)
class_legend_plot <- cowplot::ggdraw(class_legend_grob)
class_legend_plot
ggsave(file.path(dir_fig, "class_bar_213samples_LEGEND_ONLY.png"), plot = class_legend_plot, width = 10, height = 14, dpi = 300)
```

# Order and family plots
```{r}
ps_ra_order <- subset_taxa(ps_ra_top, !is.na(order))
order_levels <- get_taxa_unique(ps_ra_order, "order")
order_levels <- order_levels[!is.na(order_levels)]
palette_o <- setNames(distinctColorPalette(length(order_levels)), order_levels)
length(order_levels)

o_top_nolegend <- plot_bar(ps_ra_order, x = "Date", fill="order")+
  geom_bar(aes(fill=order), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = palette_o, na.value = "grey80") +
  scale_x_discrete(
    breaks = dates_by_site,
    labels = function(x) format(as.Date(x), "%b %Y")
  ) +
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
o_top_nolegend
ggsave(file.path(dir_fig, "order_bar_213samples_NO_LEGEND.png"), plot = o_top_nolegend, width = 26, height = 8, dpi = 300)

o_legend_only <- plot_bar(ps_ra_order, x = "Date", fill="order")+
  geom_bar(aes(fill=order), stat="identity",position="stack") +
  scale_fill_manual(values = palette_o, na.value = "grey80") +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4, title = "Order"))

order_legend_grob <- cowplot::get_legend(o_legend_only)
order_legend_plot <- cowplot::ggdraw(order_legend_grob)
order_legend_plot
ggsave(file.path(dir_fig, "order_bar_213samples_LEGEND_ONLY.png"), plot = order_legend_plot, width = 10, height = 14, dpi = 300)
```

# summer rel abund plot
```{r}
#same fix as the earlier class chunk, this pointed at ps_ra_plot which
#does not exist, should be ps_ra_top
summer <- ps_ra_top
sample_data(summer)$Date_real <- as.Date(as.character(sample_data(summer)$Date))

#only utilizing samples from summer months, july and august
keep_samples <- sample_names(summer)[format(sample_data(summer)$Date_real, "%m") %in% c("07","08")]
summer <- prune_samples(keep_samples, summer)
sample_data(summer)$Date <- droplevels(factor(sample_data(summer)$Date))

class_levels_summer <- get_taxa_unique(summer, "class")
class_levels_summer <- class_levels_summer[!is.na(class_levels_summer)]
length(class_levels_summer)

summer_dates <- levels(sample_data(summer)$Date)
summer_breaks <- summer_dates[seq(1, length(summer_dates), by = 3)]

#same legend problem as the full year class chart, 55 classes does not fit
#in one legend cleanly. same no legend plus separate legend split.
c1_summer_nolegend <- plot_bar(summer, x = "Date", fill="class") +
  geom_bar(aes(fill=class), stat="identity", position="stack") +
  theme(strip.text=element_text(face="bold", size = 16)) +
  facet_grid(. ~ Location, scales = "free", space = "free") +
  scale_fill_manual(values = euk_class_colors, na.value = "grey80") +
  scale_x_discrete(
    breaks = summer_breaks,
    labels = function(x) format(as.Date(x), "%b %Y")
  ) +
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 25)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
c1_summer_nolegend
ggsave(file.path(dir_fig, "class_summer_213samples_NO_LEGEND.png"), plot = c1_summer_nolegend, width = 20, height = 8, dpi = 300)

#reusing euk_class_colors, the full year palette, so a given class is the
#same color in both the full year and summer only charts
c1_summer_legend_only <- plot_bar(summer, x = "Date", fill="class") +
  geom_bar(aes(fill=class), stat="identity", position="stack") +
  scale_fill_manual(values = euk_class_colors, na.value = "grey80") +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4, title = "Class"))

summer_legend_grob <- cowplot::get_legend(c1_summer_legend_only)
summer_legend_plot <- cowplot::ggdraw(summer_legend_grob)
summer_legend_plot
ggsave(file.path(dir_fig, "class_summer_213samples_LEGEND_ONLY.png"), plot = summer_legend_plot, width = 10, height = 12, dpi = 300)
```

# Log transform
```{r}
#[UPDATED] ASV_mat only existed in the original load chunk, which got
#replaced with ps <- ps18S earlier. pull raw counts straight from ps instead.
otu_raw <- as(otu_table(ps), "matrix")
#taxa_are_rows = TRUE for this object, so rows are ASVs, columns are samples

relabund <- function(sample) {
  x = sample/sum(sample)
  x = x*100
  return(x)
}

#apply column wise since samples are columns here
otu_df <- apply(otu_raw, 2, relabund)
dim(otu_df)

otu_df <- otu_df[rowSums(otu_df) != 0, ]
otu_df <- otu_df[rowSums(otu_df > 0) > (0.05 * ncol(otu_df)), ]
dim(otu_df)

otu_log <- log(otu_df+1)
dim(otu_log)

otu_log2 <- vegdist(t(otu_log), "bray")
```

# Testing Normality
```{r}
library(reshape2)
asv_melted <- melt(otu_log)

qq_plot_18S <- asv_melted %>% ggplot(aes(sample = value)) +
  geom_qq() +
  geom_qq_line()
qq_plot_18S
ggsave(file.path(dir_fig, "normality_qqplot_213samples.png"), plot = qq_plot_18S, width = 8, height = 6, dpi = 300)

#hist is base R graphics not ggplot, needs the png/dev.off wrapper instead
#of ggsave
png(file.path(dir_fig, "normality_histogram_213samples.png"), width = 8, height = 6, units = "in", res = 300)
hist(as.vector(otu_log), breaks = 50, main = "Log-Transformed Data", col = "red", xlab = "log(Counts + 1)")
dev.off()
```

# NMDS
Reference: https://jkzorz.github.io/2019/06/06/NMDS.html
```{r}
set.seed(123)
otu_log_t <- t(otu_log)
bray <- vegdist(otu_log_t, "bray")
nmds <- metaMDS(bray, k = 3, trymax = 100)
nmds$stress

#[UPDATED] meta only existed in the original load chunk, replaced by ps18S.
#pull metadata for exactly the samples in the NMDS result, by name, so rows
#cannot get silently mismatched.
nmds_points <- as.data.frame(nmds$points)
nmds_samples <- rownames(nmds_points)
meta_for_nmds <- as(sample_data(ps), "data.frame")[nmds_samples, , drop = FALSE]
nmds2 <- cbind(nmds_points, meta_for_nmds)
nmds2$type <- factor(nmds2$Date)

nmds_base_plot <- ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3, "Summer" = 16, "Fall" = 5, "Winter" = 8)) +
  labs(x = "NMDS1", y = "NMDS2", shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)
nmds_base_plot

seasonplot <- ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3, "Summer" = 16, "Fall" = 5, "Winter" = 8)) +
  stat_ellipse(aes(group = Season), linetype = "dashed", size = 1) +
  labs(x = "NMDS1", y = "NMDS2", shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)
seasonplot

locationplot <- ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3, "Summer" = 16, "Fall" = 5, "Winter" = 8)) +
  stat_ellipse(aes(group = Location), linetype = "dashed", size = 1) +
  labs(x = "NMDS1", y = "NMDS2", shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)
locationplot
```

# ANOSIM test
Reference: https://jkzorz.github.io/2019/06/11/ANOSIM-test.html
```{r}
#[UPDATED] meta only existed in the original load chunk, replaced by ps18S.
#pull metadata for exactly the samples in otu_log_t, by name.
md <- as(sample_data(ps), "data.frame")[rownames(otu_log_t), , drop = FALSE]

ano_season <- anosim(otu_log_t, md$Season, distance = "bray", permutations = 9999)
ano_season

ano_site <- anosim(otu_log_t, md$Location, distance = "bray", permutations = 9999)
ano_site
```

# PERMANOVA test
```{r}
#[UPDATED] meta only existed in the original load chunk, replaced by ps18S.
metadata <- as(sample_data(ps), "data.frame")[rownames(otu_log_t), , drop = FALSE]

permanova_season <- adonis2(otu_log2 ~ Season, data = metadata, permutations = 999)
print(permanova_season)
```

# BY SEASON
```{r}
dispersion_season <- betadisper(otu_log2, metadata$Season)
disp_season_aov <- anova(dispersion_season)
disp_season_aov
```

# BY SITE
```{r}
permanova_site <- adonis2(otu_log2 ~ Location, data = metadata, permutations = 999)
print(permanova_site)

dispersion_site <- betadisper(otu_log2, metadata$Location)
disp_site_aov <- anova(dispersion_site)      
disp_site_aov  # If p < 0.05, sites have significantly different dispersion

# NOTE: temporal non-independence (repeated measures) is NOT yet handled --
# same field-wide gap Erin flagged. Restricted-permutation approach to follow.
```

# Shannon Diversity (by Season -- runs now)
```{r}
# all seasons richness
shannon_df <- estimate_richness(ps, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID")
metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))
merged_df$Location <- factor(merged_df$Location, levels = c("MIS", "STA", "MAB"))

p2 <- ggboxplot(data = merged_df,
                x = "Season",
                color = "Location",
                y = "Shannon",
                add = "jitter",
                palette = c("green3", "steelblue2", "salmon"))
p2

# per-site seasonal richness
ps_MIS <- subset_samples(ps, Location == "MIS")
ps_STA <- subset_samples(ps, Location == "STA")
ps_MAB <- subset_samples(ps, Location == "MAB")

my_comparisons <- list(c("Summer", "Spring"),
                       c("Winter", "Summer"),
                       c("Fall", "Summer"),
                       c("Fall", "Spring"),
                       c("Fall", "Winter"))

for (psx in list(MIS = ps_MIS, STA = ps_STA, MAB = ps_MAB)) {
  sdf <- estimate_richness(psx, measures = "Shannon") %>% rownames_to_column("SampleID")
  mdf <- data.frame(sample_data(psx)) %>% rownames_to_column("SampleID")
  m   <- left_join(sdf, mdf, by = "SampleID")
  m$Season <- factor(m$Season, levels = c("Winter","Spring","Summer","Fall"))
  print(
    ggboxplot(m, x = "Season", y = "Shannon", add = "jitter") +
      stat_compare_means(method = "wilcox.test", comparisons = my_comparisons,
                         label = "p.signif", hide.ns = FALSE) +
      labs(y = "Shannon Diversity Index", title = unique(m$Location)) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

# Bloom-phase diversity -- DEFERRED (no BloomStatus column yet)
```{r eval = FALSE}
# [CHANGED vs 16S] Set eval=FALSE: the metadata has no BloomStatus column
# (Pre/Bloom/Post), same data gap as 16S. Re-enable once BloomStatus is added.
# Mirrors the 16S bloom-phase Shannon plots (MIS/STA, wilcox comparisons).
```


#EXPORT CHUNK 
```{r}
# figures (objects already built by their chunks)
ggsave(file.path(dir_fig, "class_bar.png"), c1_top, width=16, height=10, dpi=300)
ggsave(file.path(dir_fig, "order_bar.png"),   o_top,        width=14, height=8, dpi=300)
ggsave(file.path(dir_fig, "family_bar.png"),  f_top,        width=14, height=9, dpi=300)
ggsave(file.path(dir_fig, "summer_class.png"),c1_summer,    width=12, height=8, dpi=300)
ggsave(file.path(dir_fig, "nmds_season.png"), seasonplot,   width=9,  height=7, dpi=300)
ggsave(file.path(dir_fig, "nmds_location.png"),locationplot,width=9,  height=7, dpi=300)

# stats -> CSV (as.data.frame makes adonis2 / anosim / anova results writable)
write.csv(as.data.frame(permanova_season), file.path(dir_stat, "permanova_season.csv"))
write.csv(as.data.frame(permanova_site),   file.path(dir_stat, "permanova_site.csv"))
write.csv(as.data.frame(disp_season_aov),  file.path(dir_stat, "betadisper_season.csv"))
write.csv(as.data.frame(disp_site_aov),    file.path(dir_stat, "betadisper_site.csv"))
# ANOSIM has no direct data.frame method -- pull the summary stats
anosim_tbl <- data.frame(
  comparison    = c("Season", "Location"),
  anosim_R      = c(ano_season$statistic, ano_site$statistic),
  significance  = c(ano_season$signif,    ano_site$signif)
)
write.csv(anosim_tbl, file.path(dir_stat, "anosim.csv"), row.names = FALSE)

# Shannon per-sample values
write.csv(merged_df[, c("SampleID","Shannon","Season","Location")],
          file.path(dir_stat, "shannon_by_sample.csv"), row.names = FALSE)

# NMDS stress (one number, worth recording per run)
writeLines(paste("NMDS stress:", round(nmds$stress, 4)),
           file.path(dir_stat, "nmds_stress.txt"))

cat("Export complete:", out_base, "\n")

```


