---
title: "16SpostProcessing"
author: "Alexander Evan Chrostowski"
date: "2026-06-05"
output: html_document
---
  
  # Loading packages
```{r}
#load needed packages
#installs are commented out, but add them back as needed
#install.packages("rlang")
packageVersion("rlang") #v1.1.0
#installed.packages("ggplot2")
library(ggplot2); packageVersion("ggplot2") #v3.4.1
theme_set(theme_bw())
#install.packages("dplyer")
library(dplyr); packageVersion("dplyr") #v1.0.8
#install.packages("tidyverse")
library(tidyverse); packageVersion("tidyverse") #v1.1.1
#for decontaminating and filtering
#BiocManager::install("decontam")
library(decontam); packageVersion("decontam") #v1.12.0
#For ordinations and other microbial community analyses
#BiocManager::install("phyloseq")
library(phyloseq); packageVersion("phyloseq")
library(vegan); packageVersion("vegan")
#BiocManager::install("corncob")
library(corncob); packageVersion("corncob"); citation("corncob")
library(reshape2); packageVersion("reshape2")
#install.packages("randomcoloR") 
library(randomcoloR); packageVersion("randomcoloR")
#install.packages("cowplot")
library(cowplot); packageVersion("cowplot")
library(grid); packageVersion("grid") #v 4.1.1
#install.packages("gridGraphics")
library(gridGraphics); packageVersion("gridGraphics") #v 0.5.1
#install.packages("janitor")
library(janitor); packageVersion("janitor") #v 2.2.0
#install.packages("ggpubr")
library(ggpubr); packageVersion("ggpubr") #v 0.5.0
library(RColorBrewer)   # for brewer.pal()
library(lubridate)      # for mdy(), year()
```

# Setting all the tables/initial variables
```{r}
#Load in the chloro/mito filtered data and separate accordingly. Entire dataset is used. 
base <- "/Users/alexchrostowski/Desktop/ALEX - Eggleston Lab /new_tables"
ASV  <- read.table(file.path(base, "ASVs_counts_ALL.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)
taxa <- as.matrix(read.table(file.path(base, "ASVs_taxonomy_ALL.csv"), sep=",", header=TRUE, row.names=1))
meta <- read.table(file.path(base, "metadata_E400to672.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)
#clean out empty columns 
meta <- meta[, colnames(meta) != "" & !is.na(colnames(meta))]

#align columns 
meta <- meta[colnames(ASV), , drop=FALSE]

#preparation for phyloseq assembly 
ASV <- otu_table(as.matrix(ASV), taxa_are_rows = TRUE)
TAX <- tax_table(taxa)
META <- sample_data(meta)

#phyloseq assembly
ps <- phyloseq(ASV, TAX, META)

#displays the corresponding table 
ps
```

# Writing and saving the tables you need for other processing
```{r}
#save the tables with no Mitochondrial/Chloroplast/Unknown sequences for use in other analyses as csv files.
write.table(as(otu_table(ps), "matrix"),"ASVsF_no_chloromito.csv",sep=",",col.names=NA)
write.table(as(tax_table(ps), "matrix"),"taxaF_no_chloromito.csv",sep=",",col.names=NA)
write.table(as(sample_data(ps), "matrix"),"metadata_no_chloromito.csv",sep=",",col.names=NA)

#save the tables with no Mitochondrial/Chloroplast/Unknown sequences for use in other analyses as csv files.
write.table(as(otu_table(ps), "matrix"), file.path(base, "ASVsF_no_chloromito.csv"), sep=",", col.names=NA)
write.table(as(tax_table(ps), "matrix"), file.path(base, "taxaF_no_chloromito.csv"), sep=",", col.names=NA)
write.table(as(sample_data(ps), "matrix"), file.path(base, "metadata_no_chloromito.csv"), sep=",", col.names=NA)
```


# Now using those tables
# most of this is just formatting things so they are in the correct data type
#re-running some things if you haven't saved your environment and are coming back to this 
```{r eval = FALSE}
# Load in the chloro/mito filtered data
ASV  <- read.table("ASVs_counts_ALL.csv", sep=",", header=TRUE, row.names=1, check.names=FALSE)
taxa <- as.matrix(read.table("ASVs_taxonomy_ALL.csv", sep=",", header=TRUE, row.names=1))
meta <- read.table("metadata_E400to672.csv", sep=",", header=TRUE, row.names=1, check.names=FALSE)

ASV = otu_table(ASV, taxa_are_rows = TRUE)
TAX = tax_table(taxa)
META = sample_data(metadata)

metadata$Bloom <- as.factor(metadata$Bloom)
str(META)

metadata %>% count(Bloom)
metadata$Bloom

metadata$Bloom[metadata$Bloom == "No"] <- 0
metadata$Bloom[metadata$Bloom == "Yes"] <- 1
metadata$Bloom <- ifelse(grepl("No", metadata$Bloom), 0, metadata$Bloom)

ps <- phyloseq(ASV, TAX, META)

metadata$BloomStatus[metadata$BloomStatus == "No"] <- NA
metadata %>% count(BloomStatus)

```


# Processing for relative abundance
```{r}
# Transform counts to relative abundance
ps_ra<-transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
sample_data(ps_ra)$Location<-factor(sample_data(ps_ra)$Location,levels=c("MIS","STA","MAB"))

# this pulls the top 0.05 instead of including all taxa -- better visualization
ps_ra_top <- filter_taxa(ps_ra,function(x) sum (x) > .005, TRUE)
ps_ra_top
```

#Handles date input as a parameter as opposed to hard coding associated dates
```{r}
sorted_dates <- format(as.Date(sample_data(ps_ra_top)$Date, format = "%m/%d/%y"), "%Y-%m-%d")
sample_data(ps_ra_top)$Date <- factor(sorted_dates, levels = sort(unique(sorted_dates)))
all_dates <- as.Date(levels(sample_data(ps_ra_top)$Date))
keep  <- !duplicated(format(all_dates, "%Y-%m"))
month_firsts <- all_dates[keep]
dates_to_show <- format(month_firsts[seq(1, length(month_firsts), by = 3)], "%Y-%m-%d")
dates_to_show
```
# PHYLUM CHART
```{r}
#Dating for the chart has been standardized
ps_phylum <- ps_ra_top

# create color scheme
phyla_colors <- c(
  "Acidobacteriota"   = "#8da0cb",   # lighter teal
  "Actinobacteriota"  = "darkolivegreen3",   # soft light blue
  "Armatimonadota"    = "#ffe066",   # soft yellow
  "Bacteroidota"      = "slategray1",   # lighter blue-gray
  "Bdellovibrionota"  = "#c48f65",   # muted tan-brown
  "Chloroflexi"        = "#7fb2b2",   # dusty teal
  "Cyanobacteria"     = "#228B22",   # keep deep green to stand out!
  "Deinococcota"      = "#e6a4b4",   # dusty rose
  "Desulfobacterota"  = "#a4c2a5",   # sage green
  "Firmicutes"        = "#f9c785",   # soft peach
  "Fusobacteriota"    = "#f4a582",   # soft orange-pink
  "Gemmatimonadota"   = "#e8b2d2",   # soft pinkish purple
  "Planctomycetota"   = "#ddadf2",   # very pale blue
  "Proteobacteria"    = "#998fc7",   # softer muted purple
  "Spirochaetota"     = "#c1e1c1",   # light green
  "Verrucomicrobiota" = "#fff6a3"    # very light yellow
)

#this is the phylum plot
p1_top <- plot_bar(ps_ra_top, x = "Date", fill="phylum")+
  geom_bar(aes(fill=phylum), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16), legend.text = element_text(size=16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = phyla_colors) +
  scale_x_discrete(
    breaks = dates_to_show,
    labels = function(x) format(as.Date(x), "%b %Y")  # Convert & format as "Jan 2024"
  ) + 
  theme(legend.position = "bottom", axis.title = element_text(size=15), title=element_text(size = 25), legend.title = element_text(size = 15)) + 
  theme(axis.text.x = element_text(size = 13, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
p1_top
dir.create("~/Desktop/16S_figures_full", showWarnings = FALSE)
ggsave("~/Desktop/16S_figures_full/phylum_bar.png", plot = p1_top, width = 14, height = 8, dpi = 300)

```


# CLASS PLOT
```{r}
#Dates have been standardized 

# create color scheme
my_colors <- c(
  "Acidimicrobiia" = "#7B68EE",     # muted teal
  "Acidobacteriae" = "#8da0cb",     # soft blue
  "Actinobacteria" = "#66c2a5",     # soft pink
  "Alphaproteobacteria" = "#1f78b4", # light green
  "Armatimonadia" = "#ffd92f",      # bright yellow
  "Bacilli" = "#e5c494",            # beige
  "Bacteroidia" = "#bfdefa",        # gray
  "Bdellovibrionia" = "#c48f65",    # muted tan-brown
  "Blastocatellia" = "#b5d8a7",     # muted sage
  "Clostridia" = "#dcb8f0",         # light lavender
  "Cyanobacteriia" = "#228B22",     # bright deep blue (pops!)
  "Deinococci" = "#e6a4b4",         # dusty rose
  "Desulfobacteria" = "#a4c2a5",    # sage green
  "Fusobacteriia" = "#fc8d62",      # orange
  "Gammaproteobacteria" = "#e78ac3",# teal again (common)
  "Gemmatimonadetes" = "#a6cee3",   # light blue
  "Holophagae" = "#b2df8a",         # light green
  "Ignavibacteria" = "#d9a066",     # warm tan
  "Kapabacteria" = "#9ecae1",       # pale steel blue
  "Leptospirae" = "#fdbf6f",        # soft orange
  "Negativicutes" = "#f2b6c6",      # pale pink
  "Oligoflexia" = "#c9c299",        # muted olive
  "OM190" = "#aad8d3",              # pale teal
  "Omnitrophia" = "#d6c6e1",        # pale purple
  "Phycisphaerae" = "#cab2d6",      # soft purple
  "Planctomycetes" = "#fb9a99",     # pinkish red
  "SL56 marine group" = "#8fd9c4",  # sea green
  "Spirochaetia" = "#c1e1c1",       # light green
  "Subgroup 22" = "#e0c28f",        # sandy tan
  "Thermoleophilia" = "#fdbf6f",    # soft yellow
  "TK10" = "#bcbddc",               # dusty periwinkle
  "vadinHA49" = "#d4a5a5",          # muted brick pink
  "Verrucomicrobiae" = "#ffff99",   # pale yellow
  "Vicinamibacteria" = "#ffed6f"    # strong yellow
)

labelling <- c(
    "Acidimicrobiia" = "Acidimicrobiia",
    "Acidobacteriae" = "Acidobacteriae",
    "Actinobacteria" = "Actinobacteria",
    "Alphaproteobacteria" = "Alphaproteobacteria",
    "Armatimonadia" = "Armatimonadia",
    "Bacilli" = "Bacilli",
    "Bacteroidia" = "Bacteroidia",
    "Bdellovibrionia" = "Bdellovibrionia",
    "Blastocatellia" = "Blastocatellia",
    "Clostridia" = "Clostridia",
    "Cyanobacteriia" = "Cyanophyceae",  # needed to rename cyanobacteria
    "Deinococci" = "Deinococci",
    "Desulfobacteria" = "Desulfobacteria",
    "Fusobacteriia" = "Fusobacteriia",
    "Gammaproteobacteria" = "Gammaproteobacteria",
    "Gemmatimonadetes" = "Gemmatimonadetes",
    "Holophagae" = "Holophagae",
    "Ignavibacteria" = "Ignavibacteria",
    "Kapabacteria" = "Kapabacteria",
    "Leptospirae" = "Leptospirae",
    "Negativicutes" = "Negativicutes",
    "Oligoflexia" = "Oligoflexia",
    "OM190" = "OM190",
    "Omnitrophia" = "Omnitrophia",
    "Phycisphaerae" = "Phycisphaerae",
    "Planctomycetes" = "Planctomycetes",
    "SL56 marine group" = "SL56 marine group",
    "Spirochaetia" = "Spirochaetia",
    "Subgroup 22" = "Subgroup 22",
    "Thermoleophilia" = "Thermoleophilia",
    "TK10" = "TK10",
    "vadinHA49" = "vadinHA49",
    "Verrucomicrobiae" = "Verrucomicrobiae",
    "Vicinamibacteria" = "Vicinamibacteria"
  )

# making the plot
c1_top <- plot_bar(subset_samples(ps_ra_top), x = "Date", fill="class")+
  geom_bar(aes(fill=class), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16), legend.text = element_text(size=16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  #scale_fill_brewer(palette = "Set3") +
  scale_fill_manual(values = my_colors, labels = labelling)+
  scale_x_discrete(
    breaks = dates_to_show,
    labels = function(x) format(as.Date(x), "%b %Y")  # Convert & format as "Jan 2024"
  ) + 
  theme(legend.position = "bottom", axis.title = element_text(size=15), title=element_text(size = 25), legend.title = element_text(size = 15)) + 
  theme(axis.text.x = element_text(size = 13, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
c1_top
ggsave("~/Desktop/16S_figures_full/class_bar.png", plot = c1_top, width = 16, height = 10, dpi = 300)
```


# Order and family plots
```{r}
order_colors <- c(
  "Acetobacterales" = "#91bfdb",   # blue
  "Armatimonadales" = "#fee08b",   # yellow
  "Bacteroidales" = "#fc8d59",     # soft red-orange
  "Burkholderiales" = "#d9ef8b",   # yellow-green
  "Caulobacterales" = "#66c2a5",   # teal
  "Cellvibrionales" = "#fdbb84",   # orange
  "Chitinophagales" = "#a6d854",   # lime green
  "Chthoniobacterales" = "#d53e4f",# strong red
  "Competibacterales" = "#b2abd2", # purple
  "Corynebacteriales" = "#998ec3", # blue-purple
  "Cyanobacteriales" = "#1a9850",  # strong green
  "Cytophagales" = "#e08214",      # rich orange
  "Diplorickettsiales" = "#fdae61",# orange-peach
  "Flavobacteriales" = "#3288bd",  # medium blue
  "Frankiales" = "slategray1",        # strong red
  "Fusobacteriales" = "#f46d43",   # coral
  "Gammaproteobacteria Incertae Sedis" = "#8c510a", # brown
  "Gemmatales" = "#66bd63",        # green
  "Gemmatimonadales" = "#a6bddb",  # pale blue
  "Holophagales" = "#d95f0e",      # burnt orange
  "Lactobacillales" = "#F8968C",   # bold red
  "Leptospirales" = "#7570b3",     # strong purple
  "Methylococcales" = "#80cdc1",   # soft teal
  "Micrococcales" = "#1f78b4",     # strong blue
  "Microtrichales" = "#cab2d6",    # light purple
  "Oceanospirillales" = "#bebada", # pastel purple
  "Opitutales" = "#fdb462",        # orange
  "Pedosphaerales" = "#ffffb3",    # pale yellow
  "Phycisphaerales" = "#b3de69",   # green
  "Pirellulales" = "#8dd3c7",      # teal
  "Pseudanabaenales" = "#66c2a5",  # teal
  "Reyranellales" = "#fccde5",     # pink
  "Rhizobiales" = "#b2df8a",       # green
  "Rhodobacterales" = "#fb8072",   # coral pink
  "SAR11 clade" = "#bc80bd",       # violet
  "Solibacterales" = "#ccebc5",    # pale green
  "Solirubrobacterales" = "#fdbf6f",# peach
  "Sphingobacteriales" = "#ff7f00",# strong orange
  "Sphingomonadales" = "#b15928",  # brown
  "Synechococcales" = "#006837",   # forest green
  "Verrucomicrobiales" = "#ffff33",# bright yellow
  "Vicinamibacterales" = "#c7e9b4",# pale green
  "Xanthomonadales" = "#fddbc7"    # light pink
)

order_levels <- get_taxa_unique(ps_ra_top, "order")
order_levels <- order_levels[!is.na(order_levels)]
missing_orders <- setdiff(order_levels, names(order_colors))
set.seed(42)  # so re-knits give the same colors for the new orders
new_order_colors <- setNames(distinctColorPalette(length(missing_orders)), missing_orders)
order_colors <- c(order_colors, new_order_colors)

#order chart with no legend
o_top_nolegend <- plot_bar(subset_samples(ps_ra_top), x = "Date", fill="order")+
  geom_bar(aes(fill=order), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = order_colors) +
  scale_x_discrete(
    breaks = dates_to_show,
    labels = function(x) format(as.Date(x), "%b %Y")
    #Convert & format as "Jan 2024"
  ) + 
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 15)) + 
  theme(axis.text.x = element_text(size = 13, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
o_top_nolegend

#order legend as its own figure
o_top_legend_only <- plot_bar(subset_samples(ps_ra_top), x = "Date", fill="order")+
  geom_bar(aes(fill=order), stat="identity",position="stack") +
  scale_fill_manual(values = order_colors) +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4))

order_legend_grob <- cowplot::get_legend(o_top_legend_only)
o_legend_plot <- cowplot::ggdraw(order_legend_grob)
o_legend_plot

#family
#standardized palette 
family_levels <- get_taxa_unique(ps_ra_top, "family")
family_levels <- family_levels[!is.na(family_levels)]
palette_f <- setNames(distinctColorPalette(length(family_levels)), family_levels)

#family chart with no legend
f_top_nolegend <- plot_bar(subset_samples(ps_ra_top), x = "Date", fill="family")+
  geom_bar(aes(fill=family), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = palette_f) +
  scale_x_discrete(
    breaks = dates_to_show,
    labels = function(x) format(as.Date(x), "%b %Y")  
    #Convert & format as "Jan 2024"
  ) + 
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 15)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
f_top_nolegend

#family legend as its own figure
f_top_legend_only <- plot_bar(subset_samples(ps_ra_top), x = "Date", fill="family")+
  geom_bar(aes(fill=family), stat="identity",position="stack") +
  scale_fill_manual(values = palette_f) +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4))

family_legend_grob <- cowplot::get_legend(f_top_legend_only)
f_legend_plot <- cowplot::ggdraw(family_legend_grob)
f_legend_plot
ggsave("~/Desktop/16S_figures_full/order_bar_NO_LEGEND.png", plot = o_top_nolegend, width = 14, height = 8, dpi = 300)
ggsave("~/Desktop/16S_figures_full/order_bar_LEGEND_ONLY.png", plot = o_legend_plot, width = 10, height = 12, dpi = 300)
ggsave("~/Desktop/16S_figures_full/family_bar_NO_LEGEND.png", plot = f_top_nolegend, width = 14, height = 8, dpi = 300)
ggsave("~/Desktop/16S_figures_full/family_bar_LEGEND_ONLY.png", plot = f_legend_plot, width = 10, height = 14, dpi = 300)
```

# summer rel abund plot
```{r}
#create summer subset
summer <- ps_ra_top
sample_data(summer)$Date_real <- as.Date(as.character(sample_data(summer)$Date))
#only utilizing samples from summer months (July & August)
summer <- subset_samples(summer, format(Date_real, "%m") %in% c("07", "08"))
sample_data(summer)$Date <- droplevels(factor(sample_data(summer)$Date))
#standardized palette 
class_levels <- get_taxa_unique(summer, "class")
class_levels <- class_levels[!is.na(class_levels)]
palette_c <- setNames(distinctColorPalette(length(class_levels)), class_levels)
summer_dates <- levels(sample_data(summer)$Date)
summer_breaks <- summer_dates[seq(1, length(summer_dates), by = 3)]  

# Plot, no legend
c1_summer_nolegend <- plot_bar(summer, x = "Date", fill="class") + 
  geom_bar(aes(fill=class), stat="identity", position="stack") + 
  theme(strip.text=element_text(face="bold", size = 16)) + 
  facet_grid(. ~ Location, scales = "free", space = "free") + 
  scale_fill_manual(values = palette_c) + 
  scale_x_discrete( 
    breaks = summer_breaks, 
    labels = function(x) format(as.Date(x), "%b %Y")
  ) +  
  theme(legend.position = "none", axis.title = element_text(size=15), title=element_text(size = 25)) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x="Sample Date", y = "Relative Abundance (%)")
c1_summer_nolegend

# Legend as its own figure
c1_summer_legend_only <- plot_bar(summer, x = "Date", fill="class") + 
  geom_bar(aes(fill=class), stat="identity", position="stack") + 
  scale_fill_manual(values = palette_c) +
  theme(legend.text = element_text(size=9), legend.title = element_text(size=12), legend.position = "right") +
  guides(fill = guide_legend(ncol = 4))

summer_legend_grob <- cowplot::get_legend(c1_summer_legend_only)
c1_summer_legend_plot <- cowplot::ggdraw(summer_legend_grob)
c1_summer_legend_plot
ggsave("~/Desktop/16S_figures_full/class_summer_NO_LEGEND.png", plot = c1_summer_nolegend, width = 14, height = 8, dpi = 300)
ggsave("~/Desktop/16S_figures_full/class_summer_LEGEND_ONLY.png", plot = c1_summer_legend_plot, width = 10, height = 12, dpi = 300)
```

# Log transform
```{r}
library(vegan)
library(phyloseq)
library(dplyr)

# pull the raw counts matrix directly from the final ps object, not the
# earlier ASV variable, which got overwritten into a phyloseq otu_table
# object several chunks ago and is no longer a plain matrix
otu_raw <- as(otu_table(ps), "matrix")
# taxa_are_rows = TRUE means rows = ASVs, columns = samples

#For Bray-Curtis, turn data into relative abundances since these data are compositional.
relabund <- function(sample) { #Write a relative abundance function
  x = sample/sum(sample)
  x = x*100
  return(x)
}

#pcoa all data bray-curtis, relative abundance
#MARGIN = 2 applies relabund per SAMPLE (columns), since rows are ASVs here
otu_df <- apply(otu_raw, 2, relabund)
dim(otu_df) #see how many ASVs are there

otu_df <- otu_df[rowSums(otu_df) != 0, ] #get rid of ASVs that have no abundance in any samples
otu_df <- otu_df[rowSums(otu_df > 0) > (0.05 * ncol(otu_df)), ]
dim(otu_df) #check to see how many ASVs are left

otu_log <- log(otu_df+1) #Log transform the relative abundance data, where I add a 1 pseudocount to everything
dim(otu_log)

#Helpful reference - http://r-sig-ecology.471788.n2.nabble.com/Variability-explanations-for-the-PCO-axes-as-in-Anderson-and-Willis-2003-td6429547.html
otu_log2 <- vegdist(t(otu_log), "bray") #calculate bray curtis dissimilarity
```

# PCoA tries -- they don't work for non-normal data though
```{r}
pcoa <- cmdscale(otu_log2, k=(3), eig=TRUE) #metric dimensional scaling (principal coordinates analysis)
barplot(pcoa$eig) #take a look at what principal coordinates best explain the data. There is a small difference between 2 and 3, so also plot principal coordinates 2 and 3.
pcoa.eig <- eigenvals(pcoa) #extract eigenvalues
##show pc explanation fig
Variance <- pcoa.eig / sum(pcoa.eig) #calculate the percent of variance explained by each axis
Variance1 <- 100 * signif(Variance[1], 2) #extract percent explained by variance 1
Variance2 <- 100 * signif(Variance[2], 2) #extract percent explained by variance 2
Variance3 <- 100 * signif(Variance[3], 3) #extract percent explained by variance 3
str(pcoa)
pcoa$points

#GGPLOT2 plotting
#pull metadata for exactly the samples in pcoa$points, by name, so rows
#cannot silently get mismatched if ordering ever differs from META
pcoa_samples <- rownames(pcoa$points)
meta_for_pcoa <- as(sample_data(ps), "data.frame")[pcoa_samples, , drop = FALSE]
pcoa2 <- cbind(pcoa$points, meta_for_pcoa) # combine the Pcoa values with metadata for easier plotting
pcoa2$type <- factor(pcoa2$Date) #adjust the levels so they plot correctly!

#principal coordinates 1 and 3
pcoa_1_3 <- pcoa2 %>% ggplot(aes(x = pcoa2[,1]*(-1), y = pcoa2[,3], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  #coord_fixed() + #aspect ratio of 1
  labs(x = paste0("PCoA1 (", Variance1, "%)"), y = paste0("PCoA3 (",Variance3, "%)"), title = "Lake Champlain Samples PCoA of Bray-Curtis Dissimilarity", shape= "Season", colour = "Location")+
  theme(legend.text.align = 0)
pcoa_1_3

# principal coordinates 1 and 2
pcoa_1_2 <- pcoa2 %>% ggplot(aes(x = pcoa2[,1]*(-1), y = pcoa2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  #coord_fixed() + #aspect ratio of 1
  labs(x = paste0("PCoA1 (", Variance1, "%)"), y = paste0("PCoA2 (",Variance2, "%)"), title = "Lake Champlain Samples PCoA of Bray-Curtis Dissimilarity", shape= "Season", colour = "Location")+
  theme(legend.text.align = 0)
pcoa_1_2

# principal coordinates 2 and 3
pcoa_2_3 <- pcoa2 %>% ggplot(aes(x = pcoa2[,2]*(-1), y = pcoa2[,3], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  #coord_fixed() + #aspect ratio of 1
  labs(x = paste0("PCoA2 (", Variance2, "%)"), y = paste0("PCoA3 (",Variance3, "%)"), title = "Lake Champlain Samples PCoA of Bray-Curtis Dissimilarity", shape= "Season", colour = "Location")+
  theme(legend.text.align = 0)
pcoa_2_3

dir.create("~/Desktop/16S_figures_full", showWarnings = FALSE)

# eigenvalue barplot needs to be captured differently since it's base R, not ggplot
png("~/Desktop/16S_figures_full/pcoa_eigenvalues.png", width = 10, height = 6, units = "in", res = 300)
barplot(pcoa$eig)
dev.off()

ggsave("~/Desktop/16S_figures_full/pcoa_axis1_axis3.png", plot = pcoa_1_3, width = 10, height = 7, dpi = 300)
ggsave("~/Desktop/16S_figures_full/pcoa_axis1_axis2.png", plot = pcoa_1_2, width = 10, height = 7, dpi = 300)
ggsave("~/Desktop/16S_figures_full/pcoa_axis2_axis3.png", plot = pcoa_2_3, width = 10, height = 7, dpi = 300)
```


# Testing Normality
```{r}
apply(otu_log, 2, function(x) shapiro.test(x)$p.value)
library(reshape2)
asv_melted <- melt(otu_log)
asv_melted%>%ggplot(aes(sample = value))+
  geom_qq()+
  geom_qq_line()
str(otu_log)

hist(as.vector(otu_log), breaks = 50, main = "Log-Transformed Data", col = "red", xlab = "log(Counts + 1)")

# Shapiro test summary (optional, but useful to have alongside the plots for context)
shapiro_pvals <- apply(otu_log, 2, function(x) shapiro.test(x)$p.value)
summary(shapiro_pvals)

# QQ plot, assigned to a variable so it's saveable
library(reshape2)
asv_melted <- melt(otu_log)
qq_plot <- asv_melted %>% ggplot(aes(sample = value)) +
  geom_qq() +
  geom_qq_line()
qq_plot

# Histogram, base R graphics so it needs the png()/dev.off() wrapper
png("~/Desktop/16S_figures_full/normality_histogram.png", width = 8, height = 6, units = "in", res = 300)
hist(as.vector(otu_log), breaks = 50, main = "Log-Transformed Data", col = "red", xlab = "log(Counts + 1)")
dev.off()

# Save the QQ plot
ggsave("~/Desktop/16S_figures_full/normality_qqplot.png", plot = qq_plot, width = 8, height = 6, dpi = 300)
```


# NDMS without the first 10 samples
Reference: https://jkzorz.github.io/2019/06/06/NMDS.html#:~:text=NMDS%20plots%20are%20non%2Dmetric,distribution%20with%20a%20long%20tail.
```{r}
# Remove first 10 samples (assuming row names are sample IDs)
otu_log_filtered <- otu_log[!rownames(otu_log) %in% as.character(400:409), ]
otu_log_filtered <- t(otu_log_filtered)

# Calculate Bray-Curtis dissimilarity
otu_log2 <- vegdist(otu_log_filtered, "bray")  

# Perform NMDS
nmds <- metaMDS(otu_log2, k = 3, trymax = 100)  

# Filter metadata to match the reduced otu_log
META <- t(META)
META_filtered <- META[, !colnames(META) %in% as.character(400:409)]

# Extract NMDS coordinates
nmds_points <- as.data.frame(nmds$points)  
nmds2 <- cbind(nmds_points, t(META_filtered))  

# Ensure type column is correctly assigned
nmds2$type <- factor(nmds2$Date)  

# PLOT THE NMDS
# can try with 2 and 3 or 1 and 3 if looks bad
ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3,  # Plus
                                "Summer" = 16,  # Unfilled Circle
                                "Fall" = 5,    # Unfilled Triangle
                                "Winter" = 8)) + # Diamond
  labs(x = "NMDS1", y = "NMDS2",  
       shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)

# drawing ellipses based on season
seasonplot <- ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3,  # Plus
                                "Summer" = 16,  # Unfilled Circle
                                "Fall" = 5,    # Unfilled Triangle
                                "Winter" = 8)) + # Diamond
  stat_ellipse(aes(group = Season), linetype = "dashed", size = 1) + 
  labs(x = "NMDS1", y = "NMDS2",  
       shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)

locationplot <- ggplot(nmds2, aes(x = nmds2[,1], y = nmds2[,2], colour = Location, shape = Season)) +
  geom_point(size = 4, stroke = 2) +
  scale_shape_manual(values = c("Spring" = 3,  # Plus
                                "Summer" = 16,  # Unfilled Circle
                                "Fall" = 5,    # Unfilled Triangle
                                "Winter" = 8)) + # Diamond
  stat_ellipse(aes(group = Location), linetype = "dashed", size = 1) + 
  labs(x = "NMDS1", y = "NMDS2",  
       shape = "Season", colour = "Location") +
  theme(legend.text.align = 0)

nmds$stress # record this and put on graph when presenting it

```

# ANOSIM test
Reference: https://jkzorz.github.io/2019/06/11/ANOSIM-test.html 
same goal as the PERMANOVA but better for this type of data
```{r}
# META_filtered comes in transposed from the NMDS chunk, variables as rows.
# MUST t() before as.data.frame or $ has no columns to find.
META_filtered <- as.data.frame(t(META_filtered))

ano <- anosim(otu_log_filtered, META_filtered$Season, distance = "bray", permutations = 9999)
ano

# for SITE
# Call:
# anosim(x = otu_log_filtered, grouping = META_filtered$Location,      permutations = 9999, distance = "bray") 
# Dissimilarity: bray 
# 
# ANOSIM statistic R: 0.1908 --> more similar, closer to 0
#       Significance: 1e-04 --> significant for site
# 
# Permutation: free
# Number of permutations: 9999


# for SEASON
# Call:
# anosim(x = otu_log_filtered, grouping = META_filtered$Season,      permutations = 9999, distance = "bray") 
# Dissimilarity: bray 
# 
# ANOSIM statistic R: 0.5549 --> closer to 1 means groups are more dissimilar
#       Significance: 1e-04 
# 
# Permutation: free
# Number of permutations: 9999

```

# PERMANOVA test
```{r}
# Convert phyloseq object to a community matrix
otu_df <- as.data.frame(otu_table(ps))
otu_df <- t(otu_df)

dim(otu_df)
otu_df <- otu_df[rowSums(otu_df) != 0, ] #get rid of ASVs that have no abundance in any samples

otu_df <- t(otu_df) # should be ASVs as rows before this step
otu_df <- otu_df[rowSums(otu_df > 0) > (0.05 * ncol(otu_df)), ]

dim(otu_df) #check to see how many ASVs are left --> [1] 839  87

otu_log <- log(otu_df+1) #Log transform the relative abundance data, where I add a 1 pseudocount to everything
dim(otu_log)
otu_log
otu_log2 <- vegdist(t(otu_log), "bray") #calculate bray curtis dissimilarity

hist(otu_log2)

metadata <- data.frame(sample_data(ps))

permanova_result <- adonis2(otu_log2 ~ Season, data = metadata, permutations = 999)
print(permanova_result)
```


# BY SEASON
```{r}

# adonis2(formula = otu_log2 ~ Season, data = metadata, permutations = 999)
#           Df SumOfSqs      R2      F Pr(>F)    
# Season    3   5.0919 0.24905 9.1755  0.001 ***
# Residual 83  15.3535 0.75095                  
# Total    86  20.4455 1.00000                  
# ---
#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

dispersion_season <- betadisper(otu_log2, metadata$Season)
anova(dispersion_season)  # If p < 0.05, seasons have significantly different dispersion

# Analysis of Variance Table
# 
# Response: Distances
#            Df  Sum Sq  Mean Sq F value   Pr(>F)   
# Groups     3 0.35827 0.119424  5.6081 0.001498 **
# Residuals 83 1.76748 0.021295                    
# ---
#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

# BY SITE
```{r}
permanova_site <- adonis2(otu_log2 ~ Location, data = metadata, permutations = 999)
print(permanova_site)
# Permutation test for adonis under reduced model
# Terms added sequentially (first to last)
# Permutation: free
# Number of permutations: 999
# 
# adonis2(formula = otu_log2 ~ Location, data = metadata, permutations = 999)
#           Df SumOfSqs      R2      F Pr(>F)   
# Location  2   1.3763 0.06732 3.0313  0.003 **
# Residual 84  19.0692 0.93268                 
# Total    86  20.4455 1.00000                 
# ---
#   Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
dispersion_site <- betadisper(otu_log2, metadata$Location)
anova(dispersion_site)  # If p < 0.05, sites have significantly different dispersion

# Analysis of Variance Table
# 
# Response: Distances
# Df  Sum Sq  Mean Sq F value Pr(>F)
# Groups     2 0.05083 0.025415  0.7487 0.4761
# Residuals 84 2.85143 0.033946     
```

# Shannon Diversity
```{r}
ps_MS <- subset_samples(ps, Location %in% c("MIS", "STA"))
meta <- data.frame(sample_data(ps_MS))
meta$Date <- as.Date(meta$Date, format = "%m/%d/%y")  # <- key line
sample_data(ps_MS) <- sample_data(meta)

ps_MS_2024 <- subset_samples(ps_MS, year(Date) == 2024)

shannon_df <- estimate_richness(ps_MS, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MS)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")

merged_df$BloomStatus <- factor(merged_df$BloomStatus, levels = c("Pre", "Bloom", "Post"))

p <- ggboxplot(data = merged_df,
               x = "BloomStatus",
               color = "Location",
               y = "Shannon",
               add = "jitter",      
               palette = c("green3", "steelblue2"))     

p

p + stat_compare_means(method = "kruskal.test") +
  stat_compare_means(comparisons = list(c("STA Pre", " STA Bloom"),
                                        c("STA Bloom", "STA Post"),
                                        c("STA Pre", "STA Post"),
                                        c("MIS Pre", " MIS Bloom"),
                                        c("MIS Bloom", "MIS Post"),
                                        c("MIS Pre", "MIS Post")),
                     method = "wilcox.test",
                     label = "p.signif")


# tests for summer vs winter

meta <- data.frame(sample_data(ps_MS))
meta$Date <- as.Date(meta$Date, format = "%m/%d/%y")  # <- key line
sample_data(ps_MS) <- sample_data(meta)

shannon_df <- estimate_richness(ps_MS, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MS)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

p <- ggboxplot(data = merged_df,
               x = "Season",
               color = "Location",
               y = "Shannon",
               add = "jitter",       
               palette = c("green3", "steelblue2"))     
p

my_comparisons <- list(c("MIS Summer", "MIS Spring"),
                       c("MIS Winter", "MIS Summer"),
                       c("STA Summer", "STA Winter"),
                       c("STA Summer", "STA Spring"))

p + stat_compare_means(method = "wilcox.test",
                       comparisons = my_comparisons,
                       label = "p.signif",
                       hide.ns = FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# all seasons richness
shannon_df <- estimate_richness(ps, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))
merged_df$Location <- factor(merged_df$Location, levels = c("MIS", "STA", "MAB"))


p2 <- ggboxplot(data = merged_df,
                x = "Season",
                color = "Location",
                y = "Shannon",
                add = "jitter",      
                palette = c("green3", "steelblue2", "salmon"))     
p2


ps_MIS <- subset_samples(ps, Location == "MIS")
ps_STA <- subset_samples(ps, Location == "STA")
ps_MAB <- subset_samples(ps, Location == "MAB")

# bloom MIS
shannon_df <- estimate_richness(ps_MIS, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MIS)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")

p5 <- ggboxplot(data = merged_df,
                x = "BloomStatus",
                color = "green3",
                y = "Shannon",
                add = "jitter",   
)     
p5

shannon_df <- estimate_richness(ps_MAB, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MAB)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")

p6 <- ggboxplot(data = merged_df,
                x = "BloomStatus",
                color = "salmon",
                y = "Shannon",
                add = "jitter",      
)      
p6

p5 + stat_compare_means(method = "wilcox.test",
                        comparisons = list(c("Pre", "Bloom"), c("Bloom", "Post"), c("Pre", "Post")),
                        label = "p.signif",
                        hide.ns = FALSE) +
  labs(y = "Shannon Diversity Index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# bloom STA
shannon_df <- estimate_richness(ps_STA, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID")
metadata_df <- data.frame(sample_data(ps_STA)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")

p6 <- ggboxplot(data = merged_df,
                x = "BloomStatus",
                color = "steelblue2",
                y = "Shannon",
                add = "jitter",        
)     
p6

p6 + stat_compare_means(method = "wilcox.test",
                        comparisons = list(c("Pre", "Bloom"), c("Bloom", "Post"), c("Pre", "Post")),
                        label = "p.signif",
                        hide.ns = FALSE) +
  labs(y = "Shannon Diversity Index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# MIS seasonal richness
shannon_df <- estimate_richness(ps_MIS, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MIS)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

p4 <- ggboxplot(data = merged_df,
                x = "Season",
                color = "green3",
                y = "Shannon",
                add = "jitter",        
)    
p4

shannon_df <- estimate_richness(ps_MAB, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_MAB)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

p7 <- ggboxplot(data = merged_df,
                x = "Season",
                color = "salmon",
                y = "Shannon",
                add = "jitter",      
)    
p7 + labs(y = "Shannon Diversity Index") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p7 + stat_compare_means(method = "wilcox.test",
                        comparisons = my_comparisons,
                        label = "p.signif",
                        hide.ns = FALSE) +
  labs(y = "Shannon Diversity Index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



# STA seasonal richness
shannon_df <- estimate_richness(ps_STA, measures = "Shannon") %>%
  rownames_to_column(var = "SampleID") 
metadata_df <- data.frame(sample_data(ps_STA)) %>%
  rownames_to_column(var = "SampleID")
merged_df <- left_join(shannon_df, metadata_df, by = "SampleID")
merged_df$Season <- factor(merged_df$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

p2 <- ggboxplot(data = merged_df,
                x = "Season",
                color = "steelblue2",
                y = "Shannon",
                add = "jitter",     
)    
p2

my_comparisons <- list(c("Summer", "Spring"),
                       c("Winter", "Summer"),
                       c("Fall", "Summer"),
                       c("Fall", "Spring"),
                       c("Fall", "Winter"))

p2 + stat_compare_means(method = "wilcox.test",
                        comparisons = my_comparisons,
                        label = "p.signif",
                        hide.ns = FALSE) +
  labs(y = "Shannon Diversity Index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Rebuild with full 251 samples (includes recovered E400-409 winter batch)
base251 <- "/Users/alexchrostowski/Desktop/16S_18S_merged_data_full/16S_251"
ASV  <- read.table(file.path(base251, "ASVs_counts_ALL.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)
taxa <- as.matrix(read.table(file.path(base251, "ASVs_taxonomy_ALL.csv"), sep=",", header=TRUE, row.names=1))
meta <- read.table(file.path(base251, "metadata_E400to672.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)

meta <- meta[, colnames(meta) != "" & !is.na(colnames(meta))]
stopifnot(all(colnames(ASV) %in% rownames(meta)))
meta <- meta[colnames(ASV), , drop=FALSE]

ASV  <- otu_table(as.matrix(ASV), taxa_are_rows = TRUE)
TAX  <- tax_table(taxa)
META <- sample_data(meta)

ps251 <- phyloseq(ASV, TAX, META)
ps251
```
# Corncob prep on the 251-sample dataset
```{r}
# Corncob setup: filter to a reasonably prevalent/abundant subset of ps251.
# Running differentialTest on all 31710 raw ASVs is both statistically weak
# (most are near zero) and very slow. Same logic as ps_ra_top, built fresh
# here so it does not depend on any earlier chunk's ps_ra/ps_ra_top state.
ps251_ra <- transform_sample_counts(ps251, function(x) x / sum(x))
ps251_filt <- filter_taxa(ps251_ra, function(x) sum(x) > .005, TRUE)
ps251_filt

# corncob wants raw counts, not relative abundance, it models counts directly
# with its own internal normalization. pull the matching raw-count taxa back
# out of ps251 using the ASV names that survived the filter above.
keep_taxa <- taxa_names(ps251_filt)
ps251_corncob <- prune_taxa(keep_taxa, ps251)
ps251_corncob
```

# 251 SAMPLE REBUILD (includes E400-409)
Rebuilding everything below against the full 251 sample dataset. E400-409
was found and processed after the original 241 sample analysis above, this
section replaces those results with the complete series. Uses _251 suffixed
variable names throughout so nothing here collides with the 241 sample
objects above.

```{r}
base251 <- "/Users/alexchrostowski/Desktop/16S_18S_merged_data_full/16S_251"
ASV  <- read.table(file.path(base251, "ASVs_counts_ALL.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)
taxa <- as.matrix(read.table(file.path(base251, "ASVs_taxonomy_ALL.csv"), sep=",", header=TRUE, row.names=1))
meta <- read.table(file.path(base251, "metadata_E400to672.csv"), sep=",", header=TRUE, row.names=1, check.names=FALSE)

meta <- meta[, colnames(meta) != "" & !is.na(colnames(meta))]
stopifnot(all(colnames(ASV) %in% rownames(meta)))
meta <- meta[colnames(ASV), , drop=FALSE]

ASV  <- otu_table(as.matrix(ASV), taxa_are_rows = TRUE)
TAX  <- tax_table(taxa)
META <- sample_data(meta)

ps251 <- phyloseq(ASV, TAX, META)
ps251
```

```{r}
ps251_ra <- transform_sample_counts(ps251, function(OTU) OTU/sum(OTU))
sample_data(ps251_ra)$Location <- factor(sample_data(ps251_ra)$Location, levels=c("MIS","STA","MAB"))
ps251_ra_top <- filter_taxa(ps251_ra, function(x) sum(x) > .005, TRUE)
ps251_ra_top
```

```{r}
sorted_dates_251 <- format(as.Date(sample_data(ps251_ra_top)$Date, format = "%m/%d/%y"), "%Y-%m-%d")
sample_data(ps251_ra_top)$Date <- factor(sorted_dates_251, levels = sort(unique(sorted_dates_251)))
all_dates_251 <- as.Date(levels(sample_data(ps251_ra_top)$Date))
keep_251 <- !duplicated(format(all_dates_251, "%Y-%m"))
month_firsts_251 <- all_dates_251[keep_251]
dates_to_show_251 <- format(month_firsts_251[seq(1, length(month_firsts_251), by = 3)], "%Y-%m-%d")
dates_to_show_251
```

```{r}
ps251_phylum <- ps251_ra_top

phyla_colors_251 <- c(
  "Acidobacteriota"   = "#8da0cb",
  "Actinobacteriota"  = "darkolivegreen3",
  "Armatimonadota"    = "#ffe066",
  "Bacteroidota"      = "slategray1",
  "Bdellovibrionota"  = "#c48f65",
  "Chloroflexi"       = "#7fb2b2",
  "Cyanobacteria"     = "#228B22",
  "Deinococcota"      = "#e6a4b4",
  "Desulfobacterota"  = "#a4c2a5",
  "Firmicutes"        = "#f9c785",
  "Fusobacteriota"    = "#f4a582",
  "Gemmatimonadota"   = "#e8b2d2",
  "Planctomycetota"   = "#ddadf2",
  "Proteobacteria"    = "#998fc7",
  "Spirochaetota"     = "#c1e1c1",
  "Verrucomicrobiota" = "#fff6a3"
)

setdiff(get_taxa_unique(ps251_ra_top, "phylum"), names(phyla_colors_251))

p1_top_251 <- plot_bar(ps251_ra_top, x = "Date", fill="phylum")+
  geom_bar(aes(fill=phylum), stat="identity",position="stack") +
  theme(strip.text=element_text(face="bold", size = 16), legend.text = element_text(size=16)) +
  facet_grid(. ~ Location, scales = "free", space = "free")+
  scale_fill_manual(values = phyla_colors_251) +
  scale_x_discrete(
    breaks = dates_to_show_251,
    labels = function(x) format(as.Date(x), "%b %Y")
  ) + 
  theme(legend.position = "bottom", axis.title = element_text(size=15), title=element_text(size = 25), legend.title = element_text(size = 15)) + 
  theme(axis.text.x = element_text(size = 13, angle = 45, hjust = 1)) +
  labs(x="Sample Date", y = "Relative Abundance (%)")
p1_top_251
ggsave("~/Desktop/16S_figures_full/phylum_bar_251.png", plot = p1_top_251, width = 14, height = 8, dpi = 300)

```


```{r}
sig_MIS_001_top <- sig_MIS_001 %>% arrange(desc(abs(estimate))) %>% head(20)
sig_STA_001_top <- sig_STA_001 %>% arrange(desc(abs(estimate))) %>% head(20)

sig_MIS_001_top$asvgenus <- factor(sig_MIS_001_top$asvgenus, levels = rev(sig_MIS_001_top$asvgenus[order(sig_MIS_001_top$estimate)]))
sig_STA_001_top$asvgenus <- factor(sig_STA_001_top$asvgenus, levels = rev(sig_STA_001_top$asvgenus[order(sig_STA_001_top$estimate)]))

palette_001 <- distinctColorPalette(length(unique(c(as.character(sig_MIS_001_top$family), as.character(sig_STA_001_top$family)))))

sig_MIS_001_plot <- ggplot(sig_MIS_001_top, aes(x = asvgenus, y = estimate, fill = family)) +
  geom_errorbar(aes(ymin = estimate - std_error, ymax = estimate + std_error), color = "black", width = .3) +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = palette_001) +
  coord_flip() +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Differential Taxa", y = "Coefficient (vs Malletts Bay)", title = "Missisquoi") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray")
sig_MIS_001_plot

sig_STA_001_plot <- ggplot(sig_STA_001_top, aes(x = asvgenus, y = estimate, fill = family)) +
  geom_errorbar(aes(ymin = estimate - std_error, ymax = estimate + std_error), color = "black", width = .3) +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = palette_001) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = " ", y = "Coefficient (vs Malletts Bay)", title = "St Albans") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray")
sig_STA_001_plot
```


```{r}
seqtab_check <- readRDS("dada2_out_paired/seqtab_nochim_ALL.rds")
tax_check <- readRDS("dada2_out_paired/taxid_ALL.rds")
check_asvs <- c("ASV_564","ASV_688","ASV_755","ASV_270","ASV_1303")
asv_to_seq <- setNames(rownames(tax_check), paste0("ASV_", seq_len(nrow(tax_check))))
nchar(asv_to_seq[check_asvs])
```

